Issues in developing multivariable molecular signatures for guiding clinical care decisions

### 2016-05-18 Documents/paper-revised.Rmd
incorporated lisas edits, waiting on final discussion revisions

---
title: "Issues in developing multivariable molecular signatures for guiding clinical care decisions"
author: "Michael C Sachs and Lisa M McShane"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: 
    pdf_document:
      citation_package: natbib
      keep_tex: yes
      fig_caption: yes
      pandoc_args: ["--bibliography", "paper.bib"]
fontsize: 12pt
geometry: margin=1in
---

@@

```{r setup, include = FALSE}
library(dplyr)
library(ggplot2)
library(knitr)
library(survival)
library(ggkm)
library(tidyr)

@@

A **biomarker signature** is a transformation of multiple individual feature measurements, typically molecular characteristics measured on a multiplex assay, to a one-dimensional space. Specifically, let $X$ denote the set of $p$ feature measurements under consideration. The signature is an unknown function $f(X): \mathbb{R}^p \mapsto \mathbb{R}^1$. The signature result may be continuous, take multiple discrete values, or be dichotomous. In principle, the signature could also depend on other clinical or pathological variables in addition to the molecular measurements, but to simplify the discussion we will focus on signatures that depend on molecular feature data only <need to do this because you refer here and later to f as having domain Rp>.  Let $S$ denote the development dataset which includes, for each of $n$ represented individuals, a feature vector $X$, an outcome $Y$, and a treatment $Z$.  $S$ is a sample of size $n$ from distribution $\mathcal{P}$ with domain $\mathcal{S}$. <Do you need this bit about domain $\mathcal{S}$? P is a distribution function so I don't think it is necessary to talk about its domain.> Let $\mathcal{F}$ be a mapping from $\mathcal{X}$ to the space of continuous functions, $\mathcal{D}$, with domain $\mathbb{R}^p$ and range $\mathbb{R}$. <The symbol $\mathcal(X) has not been defined.  It is actually the space of development data sets that include not only X but also Y and Z for each of n individuals.  Please use a letter different than X or readers be confused and think that it includes only the feature vectors.> <Here again you describe f as having domain Rp.>  Thus $\mathcal{F}: \mathcal{X} \mapsto \mathcal{D}$ denotes the process or algorithm through which a particular $f$ is estimated. We do not place any other restrictions on $\mathcal{F}$, it could be a clustering approach, a regression approach, a combination of both, or something else entirely. We will use $\mathcal{F}$ to denote the manner in which $f$ is estimated and will write $f \in \mathcal{F}$ to denote that $f$ is estimated with the class of methods $\mathcal{F}$.

Let $\phi: \mathcal{D} \times \mathcal{X} \mapsto \mathbb{R}$ <Change mathcal(X) here if you decide to follow my suggestion above to change X to a different letter.> denote the statistic that quantifies the performance of the function $f$, such as predictive accuracy, mean squared error, or area under the receiver operating characteristic (ROC) curve (AUC). This could also be a measure of association, such as an odds ratio, hazard ratio, or log-rank statistic. This is a function of both $f$ and $S$.  We are interested in estimating $E_\mathcal{P}[\phi_{f^*}(S)]$, which is the expected error under the data generation mechanism (distribution $\mathcal{P}$), for a particular $f^* \in \mathcal{F}$. This allows us to understand how the signature will perform on future observations generated from $\mathcal{P}$. We may also be interested in estimating $E_\mathcal{P}[\phi_f(S)]$ *for all* $f \in \mathcal{F}$, which is the generalization performance for $f$ generated using mechanism $\mathcal{F}$. This doesn't guide outside researchers as to which specific $f$ to use, yet it is useful for development because it tells us how much signal is in the data. As shorthand we will write this as $E_\mathcal{P}[\phi_\mathcal{F}(S)]$

@@

The inherent statistical challenge is how to both develop a signature $f \in \mathcal{F}$ and obtain a valid assessment of its performance.  Additionally, one should provide a specification of $f$ for others to use. Typically, a specific $f$ is estimated using $\mathcal{F}$ based on some training data. The class of methods $\mathcal{F}$ can comprise a variety of different computational approaches. In recent years there has been an explosion in the literature of computational approaches to classification and prediction, and we do not intend to summarize them all here. Some excellent reviews are provided by @hastie2009elements and @moons2012riskI. The main considerations in signature estimation are identifying the features to include, deciding what transformations to apply to the feature measurements, determining how to combine the feature measurements, and determining what, if any, thresholds/cutoffs should be applied to the resulting signature value.

Our focus here will be mainly on the development of signatures using supervised methods, meaning methods that use information on an outcome variable it is desired to predict during the signature development process.  Regression methods are a commonly used class of supervised methods.  Oncotype DX [@paik2004multigene] is an example of a prognostic signature used for clinical decision making in early stage breast cancer that was developed using supervised methods.  In the case of Oncotype DX, the outcome $Y$ was time to distant recurrence of breast cancer and the feature data were gene expression values measured in breast tumor specimens.   An common supervised approach to identifying treatment-selection signatures is to use regression techniques to estimate a signature that has a strong interaction with a particular treatment. Supervised methods are contrasted with unsupervised methods which use only the feature data to derive a signature.  An example of a signature developed by unsupervised methods is the signature that identifies biological subclasses of diffuse large B-cell lymphoma, which were originally developed using clustering methods and subsequently were found to be associated with prognosis [@alizadeh2000lymphoma]. It is possible, and quite common in high-dimensional data settings, to combine multiple approaches when estimated a signature. For instance, a data-reduction step by variable selection or clustering may be performed before doing regression analysis on the resulting components. A review by Subramanian and Simon identified a large collection of gene expression–based prognostic signatures in lung cancer which had been developed using a wide variety of methods [@subramanian2010lung].

The focus of this paper is how to obtain a valid estimate of signature performance as quantified by a metric $phi$, that is, a good estimate of $E_\mathcal{P}[\phi_\mathcal{F}(S)]$. The principles discussed apply no matter what signature development method is used.  Performance depends on the true signal in the data and the specific algorithm $\mathcal{F}$ used for development.  The performance will reflect that expected of signatures developed applying the prescribed methods to data drawn from the distribution $mathcal{P}$.  For signature deployment, one would also want a particular specification of $f$ for others to use on independent data (e.g., in clinical practice). It is not trivial to conduct this evaluation in a valid manner when there is a limited data set available to both define a signature and assess its performance.  We illustrate the potential for bias to creep into the performance and review some strategies to avoid the pitfalls that lead to these biases.  We evaluate various strategies through a limited simulation study and compare the performance of some of those strategies using a real data exmaple involving the developmnet of a signature for use in making treatment decisions for patients with lung cancerseries.

@@

Typically a performance metric $\phi$ used during signature development is degined to quantify either calibration or discrimination, or some combination of the two. <Define calibration and discrimination.> Calibration assesses <fill in>.  To assess discrimination, a different metric $\phi$ may be used, such as the area under the ROC curve. Measures of association such as the odds ratio, hazard ratio, or difference in survival probabilities are also commonly used performance metrics, although their value for biomarker signatures has been debated [@pepe2004limitations]. In @zhu2010prognostic, $\phi$ was the hazard ratio comparing the high risk and low risk groups. The risk groups were determined by the signature $f$, which was estimated using the JBL.10 observation arm data. The signature was then applied to the same data that were used to build it to define risk groups, producing a the hazard ratio estimate of 15 for a prognostic effect in the observation arm. Subsequent evaluations of the signature's performance on independent data resulted in hazard ratio estimates in the 2 - 3 range, substantially smaller than the original estimate of 15. Next we reanalyze the JBR.10 data and illustrate some methods to avoid bias in evaluation of signature performance.

@@

A traditional method to avoid overfitting is the split sample approach. First, randomly partition $S$ into the training sample $S_t$ and the holdout sample $S_h$ with sample sizes $n_t$ and $n_h$, respectively. Then, $S_h$ is hidden from the analyst while $\mathcal{F}$ is applied to $S_t$ to estimate the signature function $f_t$. For fixed $f_t$, $\hat{E}[\phi_{f_t}(S_h)]$ is an unbiased estimator of $E_{\mathcal{P}}[\phi_{f_t}(S)]$ <Is this notation correct?> but it may underestimate $E_\mathcal{P}[\phi_\mathcal{F}(S)]$ which refers to a predictors built from a sample of size n. @dobbin2011optimally investigate how to optimally split a dataset into training and holdout partitions. As n, $n_t$, and $n_h$ all become very large the degree of underestimation will become negligible as the performance of all predcitors will approach that of a signature developed from a sample of infinite size from the distribution mathcal{P}.  The specific form of $f_t$ that is fixed using $S_t$ can be reported as the function for others to use, therefore the aforementioned estimator applies for that specific $f_t$. The drawback of the split-sample approach is that the performance of $f_t$ is likely to be inferior to the performance of a function $f^*$ derived using the same approach applied to the entire data set. @dobbin2011predictorsamplesize investigate how to choose the sample size n so that there is high probability that the signature developed will have expected performance within some specified margin of the the optimal signature that could be developed if the development data set had infinite sample size.

Another approach to avoid overfitting is cross-validation, which is a resampling based approach. For a fixed integer $k$, which can be between 1 and $n$, we randomly partition the full data set $S$ into subgroups of size $k$ (assume for simplicity here that n is evenly divisible by k so that n/k subgroups are formed). For each $k$, $f_{-k}$ is estimated and fixed by appling $\mathcal{F}$ on $S_{-k}$ which is the subset of $S$ that is disjoint from $S_k$. Then, we get an estimate $\hat{E}[\phi_{f_{-k}}(S_k)]$ which is an estimate of $E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]$, but it will have some pessimistic bias because a signature developed from a sample of size n-k has less information to work with than one based on a sample of size n.  When n is very large relative to k this bias is expected to be minimal.  This process is repeated $K$ times to yield $K$ estimates and then these are averaged to obtain the performance estimate. <Did you just cycled through the K subgroups in the partition or did you randomly draw subsets?  Hopefully it is the former, in which case K = n/k. This should be clarified in the text.> This process is called "leave $k$ out" or "$n/k$ fold" cross-validation. Note that for each subgroup in the partition, we obtain a new estimate of $f$, therefore we are only estimating $E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]$ and not the performance for a single specified signature. Typically, if a specific form for $f$ is desired, it would be estimated using the entire dataset $S$ to yield $f^*$ as above. Importantly one should report the cross-validated estimate just described and not the naive esitimate that would be obtained by substituting the full data S into the performance metric calculation for the signature $f^*$.  The problematic aspects of such naive "resubstitution estimates" of performance are discuss in more detail below.

@@

Another variation on all of these methods is the concept of pre-validation [@tibshirani2002pre]. With pre-validation, instead of computing the statistic $\phi$ for each of the held-out subsets ($S_{-b}$ for the bootstrap or $S_{k}$ for cross-validation), the fitted signature $\hat{f}(X_i)$ is estimated for $X_i \in S_{-b}$ where $\hat{f}$ is estimated using $S_{b}$. This process is repeated to obtain a set of pre-validated signature estimates $\hat{f}$ which are then used to calculate $\phi$ <I think that we have been using phi as the function that defines the performance metric rather than the performance metric estimate, correct?>. For single-step holdout <what is this?>, this process is equivalent to what is described above. For cross-validation and the bootstrap, this process avoids the problem of having too few cases to estimate the statistic $\phi$ on the held-out dataset. <Do you mean that this avoids the diffciulties in using cross-validation adn bootstrapping?>

@@

We compare each of the methods described above: split-sample holdout, cross-validation, bootstrap, and pre-validation, along with several commonly used but biased approaches. The biased approaches are described in Table \ref{descript}. Two of the biased approaches use the full sample to select the features, followed by fitting the multivariable model on the holdout subset. This is referred to as "parital holdout" or "partial CV" when using split-sample holdout or cross-validation as the validation step, respectively. We also implemented the naive resubstitution approach, wherein the model is trained and evaluated on the same dataset, and the partial resubstitution approach wherein the model is developed on a training subset but then evaluated on the combined training adn holdout data sets. Our main interest is in comparing the bias and variance of the resulting estimates of $E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]$. In our simulation, we look at two different performance metrics, the area under the ROC curve (AUC) and the odds ratio for the outcome comparing the signature groups.

@@

Not surprisingly, the resubstitution estimates are optimistically biased: the naive resubstitution estimate of the AUC is 44% larger than it should be and the <log?> OR estimate is over 2 times <that is on the log scale but the table is not log scale> higher than it should be, on average. Partial resubstitution, partial holdout, and partial cross-validation estimates do not ameliorate the bias very much. Investigators are tempted to use partial holdout estimates as it is more convneient to not have to carry along a large number of feature measurements into the modeling part of the signature development process, and they may think that the resulting perofrmance esitmates are still close to valid, as only half of the data are used to form the estimates;  However here we see that these versions are still severely biased and should not be reported as valid assessments of the performance of biomarker signatures.

The split-sample holdout, cross-validation, bootstrap, and pre-validation methods all produce esentially unbiased estimates in the simulated example, with their mean AUCs being nearly 0.5 and the mean ORs being nearly 1. We can compare the spread of the distributions to get a sense of the differences in precision of the estimates. The bootstrap approach appears to be the most precise, followed by the cross-validation, holdout, and finally the pre-validation. The bootstrap, as intended, is a more efficient, smoothed version of the cross validation estimate (Figure \ref{fig1}, \ref{fig2}). It provides the best balance between allocating data to <I still don't like the word "precisely" so I have deleted it. Could say "adequately."  Precisely gives the impression that there is a single underlying form of the predictor and it is just a matter of more precisely fitting model coefficients.> train the signature and having independent data remaining to precisely estimate the statistic $\phi$.

@@

After processing the data as described above, we performed a gene selection step wherein we fit univariate Cox regression models with disease specific survival as the outcome and each gene as the single predictor. Genes with univariate p-values less than 0.005 were preliminarily selected for further analysis. Then, each gene from the preliminary list was weighted by its univariate Cox regression coefficient, and the resulting weighted gene expression values were summed to form risk scores. Genes were selected for inclusion in the risk score in a forward selection manner.  Starting with the most significant weighted gene, the gene that when added to the risk score improved the concordance between survival times and the risk score most was selected next. If no gene improved the concordance, the process was stopped. All genes on the final selected list were included in a multivariable Cox regression model to fit the final risk score <Did you throw away the original coefficients and refit all of them in the final model that included the genes on the final list?>. The cutoff applied to the continuous risk score that split the patients into two risk groups corresponding to the smallest log-rank statistic p-value was selected. 
The signature development procedure was described in the introduction. There is both a feature selection step, and a multivariable estimation step. This results in a continuous signature which is the linear predictor of a Cox regression model. The signature is dichotomized by selecting the cutoff that yields the most significant log-rank statistic for comparing the resulting risk groups. Discrimination of the signature is assessed using the concordance statistic as implemented in the \texttt{survival} package in \texttt{R} [@survival]. To paraphrase the help file: this is defined as the probability of agreement for any two randomly chosen observations, which in this case means that the observation with the shorter survival time also has the larger signature value. This is similar to an interpretation of the AUC for binary data.

@@

We also show plots to assess the ability of the signature to be useful for treatment selection (Figure \ref{fig3}). These plots show the survival curves comparing treatment arms grouped in panels by the risk score. As described by @polley2013statistical, the idea is to determine whether the treatment is beneficial in one group and not benficial or harmful in another group, indicating that different treatment decisions would be made based on the signature. On one hand, the overfit signature shows dramatic differences in treatment efficacy between the low risk and high risk groups. In fact it appears that the treatment is harmful in the low risk group, but highly beneficial in the high risk group. The prevalidated signature on the other hand, shows differences that are much less dramatic. It appears that the treatment is mildly beneficial in both groups, possibly to a higher degree in the high risk group. This suggests that the dramatic predictive value of the signature was merely an artifact of overfitting the signature to the OBS arm data.  @simon2011re explain the flaws in this type of approach in which one attempts to develop a signature of differential treatment benefit ("predictive") by developing a prognostic signature on data from the control arm of a trial and then applying that signature to both the control and experimental arm data to establish predictiveness.

In a multivariable Cox model we observe similar trends when comparing the prevalidated signature to the overfit signature. We fit two regression models. In the first, the aim is to assess the prognostic value of the signature by estimating the hazard ratio for the high risk versus low risk groups, adjusted for tumor histologic subtype, stage, age, and sex. In the second model, the idea is to assess the predictive value of the signature by estimating the treatment by signature interaction effect, adjusting for the same clinical covariates. The results are reported in Table \ref{adjhr}. For the partial resubstitution approach, we find an extreme hazard ratio of nearly 40 for the prognostic effect, and a strong and significant treatment by signature interaction. Using the prevalidated signature, the effect estimates are much smaller and of small magnitude in comparison to the standard clinical features. <You shoudl present a tabel witht the ful lresults of the mutivariable Cox regression fits.> Note that the inference from the pre-validated model is not exactly correct either, because the resampling procedure induces a correlation among the observations (the prevalidated signature values are estimated from heavily overlapping data). Despite this, these results are unimpressive and would not be considered promising for clinical use.

@@

@@

I STOPPED HERE AND WILL HAVE TO WORK MORE ON THE DISCUSSION TOMORROW.

@@

All analysis code and the source files for this manuscript is available from the authors' webpage.

>>>>

---
title: "Issues in developing multivariable molecular signatures for guiding clinical care decisions"
author: "Michael C Sachs and Lisa M McShane"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: 
    pdf_document:
      citation_package: natbib
      keep_tex: yes
      fig_caption: yes
      pandoc_args: ["--bibliography", "paper.bib"]
header-includes:
    - \usepackage{setspace}
    - \doublespacing
    - \usepackage{lineno}
    - \linenumbers
fontsize: 11pt
geometry: margin=1in
---

@@

```{r setup, include = FALSE}
library(dplyr)
library(ggplot2)
library(knitr)
library(survival)
library(ggkm)
library(tidyr)
library(stargazer)

@@

A **biomarker signature** is a transformation of multiple individual feature measurements, typically molecular characteristics measured on a multiplex assay, to a one-dimensional space. Specifically, let $X$ denote the set of $p$ feature measurements under consideration. The signature is an unknown function $f(X): \mathbb{R}^p \mapsto \mathbb{R}^1$. The signature result may be continuous, take multiple discrete values, or be dichotomous. In principle, the signature could also depend on other clinical or pathological variables in addition to the molecular measurements, but to simplify the discussion we will focus on signatures that depend on molecular feature data only.  Let $S$ denote the development dataset which includes, for each of $n$ represented individuals, a feature vector $X$, an outcome $Y$, and a treatment $Z$.  $S$ is a sample of size $n$ from distribution $\mathcal{P}$ with domain $\mathcal{S}$. Let $\mathcal{F}$ be a mapping from $\mathcal{S}$ to the space of continuous functions, $\mathcal{D}$, with domain $\mathbb{R}^p$ and range $\mathbb{R}$. Thus $\mathcal{F}: \mathcal{S} \mapsto \mathcal{D}$ denotes the process or algorithm through which a particular $f$ is estimated. We do not place any other restrictions on $\mathcal{F}$, it could be a clustering approach, a regression approach, a combination of both, or something else entirely. We will use $\mathcal{F}$ to denote the manner in which $f$ is estimated and will write $f \in \mathcal{F}$ to denote that $f$ is estimated with the class of methods $\mathcal{F}$.

Let $\phi: \mathcal{D} \times \mathcal{S} \mapsto \mathbb{R}$ denote the statistic that quantifies the performance of the function $f$, such as predictive accuracy, mean squared error, or area under the receiver operating characteristic (ROC) curve (AUC). This could also be a measure of association, such as an odds ratio, hazard ratio, or log-rank statistic. This is a function of both $f$ and $S$.  We are interested in estimating $E_\mathcal{P}[\phi_{f^*}(S)]$, which is the expected error under the data generation mechanism (distribution $\mathcal{P}$), for a particular $f^* \in \mathcal{F}$. This allows us to understand how the signature will perform on future observations generated from $\mathcal{P}$. We may also be interested in estimating $E_\mathcal{P}[\phi_f(S)]$ **for all** $f \in \mathcal{F}$, which is the generalization performance for $f$ generated using mechanism $\mathcal{F}$. This doesn't guide outside researchers as to which specific $f$ to use, yet it is useful for development because it tells us how much signal is in the data. As shorthand we will write this as $E_\mathcal{P}[\phi_\mathcal{F}(S)]$.

@@

The inherent statistical challenge is how to both develop a signature $f \in \mathcal{F}$ and obtain a valid assessment of its performance.  Additionally, one should provide a specification of $f$ for others to use. Typically, a specific $f$ is estimated using $\mathcal{F}$ based on some training data. The class of methods $\mathcal{F}$ can comprise a variety of different computational approaches. In recent years there has been an explosion in the literature of computational approaches to classification and prediction, and we do not intend to summarize them all here. Some excellent reviews are provided by @hastie2009elements and @moons2012riskI. The main considerations in signature estimation are identifying the features to include, deciding what transformations to apply to the feature measurements, determining how to combine the feature measurements, and determining what, if any, thresholds or cutoffs should be applied to the resulting signature value.

Our focus here will be mainly on the development of signatures using supervised methods, meaning methods that use information on an outcome variable it is desired to predict during the signature development process.  Regression methods are a commonly used class of supervised methods.  Oncotype DX [@paik2004multigene] is an example of a prognostic signature used for clinical decision making in early stage breast cancer that was developed using supervised methods.  In the case of Oncotype DX, the outcome $Y$ was time to distant recurrence of breast cancer and the feature data were gene expression values measured in breast tumor specimens.  A common supervised approach to identifying treatment-selection signatures is to use regression techniques to estimate a signature that has a strong interaction with a particular treatment. Supervised methods are contrasted with unsupervised methods which use only the feature data to derive a signature.  An example of a signature developed by unsupervised methods is the signature that identifies biological subclasses of diffuse large B-cell lymphoma, which were originally developed using clustering methods and subsequently were found to be associated with prognosis [@alizadeh2000distinct]. It is possible, and quite common in high-dimensional data settings, to combine multiple approaches when estimated a signature. For instance, a data-reduction step by variable selection or clustering may be performed before doing regression analysis on the resulting components. A review by Subramanian and Simon identified a large collection of gene expression–based prognostic signatures in lung cancer which had been developed using a wide variety of methods [@subramanian2010gene].

The focus of this paper is how to obtain a valid estimate of signature performance as quantified by a metric $phi$, that is, a good estimate of $E_\mathcal{P}[\phi_\mathcal{F}(S)]$. The principles discussed apply no matter what signature development method is used.  Performance depends on the true signal in the data and the specific algorithm $\mathcal{F}$ used for development.  The performance will reflect that expected of signatures developed applying the prescribed methods to data drawn from the distribution $\mathcal{P}$.  For signature deployment, one would also want a particular specification of $f$ for others to use on independent data (e.g., in clinical practice). It is not trivial to conduct this evaluation in a valid manner when there is a limited data set available to both define a signature and assess its performance.  We illustrate the potential for bias to creep into the performance estimation and review some strategies to avoid the pitfalls that lead to these biases.  We evaluate various strategies through a simulation study and compare the performance of some of those strategies using a real data exmaple involving the development of a signature for use in making treatment decisions for patients with lung cancer.

@@

Typically a performance metric $\phi$ used during signature development is degined to quantify either calibration or discrimination, or some combination of the two. Calibration assesses how well the signature based predictions compare to the observed outcomes. Discrimination assesses how well the signature distinguishes between groups of patients who do or do not experience an event.  To assess discrimination, a different metric $\phi$ may be used, such as the area under the ROC curve. Measures of association such as the odds ratio, hazard ratio, or difference in survival probabilities are also commonly used performance metrics, although their value for biomarker signatures has been debated [@pepe2004limitations]. In @zhu2010prognostic, $\phi$ was the hazard ratio comparing the high risk and low risk groups. The risk groups were determined by the signature $f$, which was estimated using the JBL.10 observation arm data. The signature was then applied to the same data that were used to build it to define risk groups, producing a the hazard ratio estimate of 15 for a prognostic effect in the observation arm. Subsequent evaluations of the signature's performance on independent data resulted in hazard ratio estimates in the 2 - 3 range, substantially smaller than the original estimate of 15. Next we reanalyze the JBR.10 data and illustrate some methods to avoid bias in evaluation of signature performance.

@@

A traditional method to avoid overfitting is the split sample approach. First, randomly partition $S$ into the training sample $S_t$ and the holdout sample $S_h$ with sample sizes $n_t$ and $n_h$, respectively. Then, $S_h$ is hidden from the analyst while $\mathcal{F}$ is applied to $S_t$ to estimate the signature function $f_t$. For fixed $f_t$, $\hat{E}[\phi_{f_t}(S_h)]$ is an unbiased estimator of $E_{\mathcal{P}}[\phi_{f_t}(S)]$. @dobbin2011optimally investigate how to optimally split a dataset into training and holdout partitions. The specific form of $f_t$ that is fixed using $S_t$ can be reported as the function for others to use, therefore the aforementioned estimator applies for that specific $f_t$. The drawback of the split-sample approach is that the performance of $f_t$ is likely to be inferior to the performance of a function $f^*$ derived using the same approach applied to the entire data set. @dobbin2007sample investigate how to choose the sample size n so that there is high probability that the signature developed will have expected performance within some specified margin of the the optimal signature that could be developed if the development data set had infinite sample size.

Another approach to avoid overfitting is cross-validation, which is a resampling based approach. For a fixed integer $k$, which can be between 1 and $n$, we randomly partition the full data set $S$ into subgroups of size $k$ (assume for simplicity here that $n$ is evenly divisible by $k$ so that $n/k$ subgroups are formed). For each $k$, $f_{-k}$ is estimated and fixed by appling $\mathcal{F}$ on $S_{-k}$ which is the subset of $S$ that is disjoint from $S_k$. Then, we get an estimate $\hat{E}[\phi_{f_{-k}}(S_k)]$ which is an estimate of $E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]$. This process is repeated $K$ times to yield $K$ estimates and then these are averaged to obtain the performance estimate. This process is called "leave $k$ out" or "$n/k$ fold" cross-validation. Note that for each subgroup in the partition, we obtain a new estimate of $f$, therefore we are only estimating $E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]$ and not the performance for a single specified signature. Typically, if a specific form for $f$ is desired, it would be estimated using the entire dataset $S$ to yield $f^*$ as above. Importantly one should report the cross-validated estimate just described and not the naive esitimate that would be obtained by substituting the full data $S$ into the performance metric calculation for the signature $f^*$.  The problematic aspects of such naive "resubstitution estimates" of performance are discuss in more detail below.

@@

Another variation on all of these methods is the concept of pre-validation [@tibshirani2002pre]. With pre-validation, instead of computing the statistic $\phi$ for each of the held-out subsets ($S_{-b}$ for the bootstrap or $S_{k}$ for cross-validation), the fitted signature $\hat{f}(X_i)$ is estimated for $X_i \in S_{-b}$ where $\hat{f}$ is estimated using $S_{b}$. This process is repeated to obtain a set of pre-validated signature estimates $\hat{f}$ which are then used to calculate $\phi$. For single-step split sample training and validation, this process is equivalent to what is described above. For cross-validation and the bootstrap, this process avoids the problem of having too few cases to estimate the statistic $\phi$ on each of the smaller held-out datasets.

@@

We compare each of the methods described above: split-sample holdout, cross-validation, bootstrap, and pre-validation, along with several commonly used but biased approaches. The biased approaches are described in Table \ref{descript}. Two of the biased approaches use the full sample to select the features, followed by fitting the multivariable model on the holdout subset. This is referred to as "parital holdout" or "partial CV" when using split-sample holdout or cross-validation as the validation step, respectively. We also implemented the naïve resubstitution approach, wherein the model is trained and evaluated on the same dataset, and the partial resubstitution approach wherein the model is developed on a training subset but then evaluated on the combined training adn holdout data sets. Our main interest is in comparing the bias and variance of the resulting estimates of $E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]$. In our simulation, we look at two different performance metrics, the area under the ROC curve (AUC) and the odds ratio for the outcome comparing the signature groups.

@@

Not surprisingly, the resubstitution estimates are optimistically biased: the naive resubstitution estimate of the AUC is 44% larger than it should be and the log OR estimate is over 2 times higher than it should be, on average. Partial resubstitution, partial holdout, and partial cross-validation estimates do not ameliorate the bias very much. Investigators are tempted to use partial holdout estimates as it is more convenient to not have to carry along a large number of feature measurements into the modeling part of the signature development process, and they may think that the resulting performance estimates are still close to valid, as only half of the data are used to form the estimates.  However, here we see that these versions are still severely biased and should not be reported as valid assessments of the performance of biomarker signatures.

The split-sample holdout, cross-validation, bootstrap, and pre-validation methods all produce esentially unbiased estimates in the simulated example, with their mean AUCs being nearly 0.5 and the mean ORs being nearly 1. We can compare the spread of the distributions to get a sense of the differences in precision of the estimates. The bootstrap approach appears to be the most precise, followed by the cross-validation, holdout, and finally the pre-validation. The bootstrap, as intended, is a more efficient, smoothed version of the cross validation estimate (Figure \ref{fig1}, \ref{fig2}). It provides the best balance between allocating data to train the signature and having independent data remaining to precisely estimate the statistic $\phi$.

@@

After processing the data as described above, we performed a gene selection step wherein we fit univariate Cox regression models with disease specific survival as the outcome and each gene as the single predictor. Genes with univariate p-values less than 0.005 were preliminarily selected for further analysis. Then, each gene from the preliminary list was weighted by its univariate Cox regression coefficient, and the resulting weighted gene expression values were summed to form risk scores. Genes were selected for inclusion in the risk score in a forward selection manner.  Starting with the most significant weighted gene, the gene that when added to the risk score improved the concordance between survival times and the risk score most was selected next. If no gene improved the concordance, the process was stopped. All genes on the final selected list were included in a multivariable Cox regression model to fit the final risk score with new coefficients. We selected the cutoff that split the patients into two risk groups corresponding to the smallest log-rank statistic p-value when applied to the continuous risk score.

The signature development procedure was described in the introduction. There is both a feature selection step, and a multivariable estimation step. This results in a continuous signature which is the linear predictor of a Cox regression model. The signature is dichotomized by selecting the cutoff that yields the most significant log-rank statistic for comparing the resulting risk groups. Discrimination of the signature is assessed using the concordance statistic as implemented in the \texttt{survival} package in \texttt{R} [@survival]. To paraphrase the help file: this is defined as the probability of agreement for any two randomly chosen observations, which in this case means that the observation with the shorter survival time also has the larger signature value. This is similar to an interpretation of the AUC for binary data.

@@

We also show plots to assess the ability of the signature to be useful for treatment selection (Figure \ref{fig3}). These plots show the survival curves comparing treatment arms grouped in panels by the risk score. As described by @polley2013statistical, the idea is to determine whether the treatment is beneficial in one group and not benficial or harmful in another group, indicating that different treatment decisions would be made based on the signature. On one hand, the overfit signature shows dramatic differences in treatment efficacy between the low risk and high risk groups. In fact it appears that the treatment is harmful in the low risk group, but highly beneficial in the high risk group. The prevalidated signature on the other hand, shows differences that are much less dramatic. It appears that the treatment is mildly beneficial in both groups, possibly to a higher degree in the high risk group. This suggests that the dramatic predictive value of the signature was merely an artifact of overfitting the signature to the OBS arm data.  @simon2011re explain the flaws in this type of approach in which one attempts to develop a signature for treatment selection by developing a prognostic signature on data from the control arm of a trial and then applying that signature to both the control and experimental arm data to establish predictiveness.

In a multivariable Cox model we observe similar trends when comparing the prevalidated signature to the overfit signature. We fit two regression models. In the first, the aim is to assess the prognostic value of the signature by estimating the hazard ratio for the high risk versus low risk groups, adjusted for tumor histologic subtype, stage, age, and sex. In the second model, the idea is to assess the predictive value of the signature by estimating the treatment by signature interaction effect, adjusting for the same clinical covariates. The results are reported in Table \ref{adjhr}. The full model summaries are report in Table \ref{multimodel}. For the partial resubstitution approach, we find an extreme hazard ratio of nearly 40 for the prognostic effect, and a strong and significant treatment by signature interaction. Using the prevalidated signature, the effect estimates are much smaller and of small magnitude in comparison to the standard clinical features. Note that the inference from the pre-validated model is not exactly correct either, because the resampling procedure induces a correlation among the observations (the prevalidated signature values are estimated from heavily overlapping data). Despite this, these results are unimpressive and would not be considered promising for clinical use.

@@

```{r modtab, results = "asis"}
stargazer(fit.train, fit.preval, fit.train.inter, fit.preval.inter, 
          column.labels = rep(c("Resubstitution", "Prevalidation"), 2), 
          covariate.labels = c("High Risk", "Trt: OBS", "Male Gender", "Stage > I", "Age in Years", "Histology: LCUC", 
                               "Histology: SQCC", "High Risk * Trt:OBS"), 
          dep.var.labels = "Disease-specific Survival", ci = TRUE, ci.separator = " to ", digits = 1, 
          keep.stat = c("n", "logrank"), header = FALSE, apply.coef = exp,  apply.ci = exp, p.auto = FALSE,
          ci.custom = lapply(list(fit.train, fit.preval, fit.train.inter, fit.preval.inter), confint),
          title = "Complete Results of Multivariable Regression Models. Entries are hazard ratios and 95 percent confidence intervals.  \\label{multimodel}")
```

@@

@@

All analysis code and the source files for this manuscript are available from the authors' webpage.
